MLH1 Results Outline

### 2020-05-01 doc/Results_v2.rmd
updating the figures

---
title: "MLH1 Results Outline"
date: 'null'
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '5'
bibliography: refs_Main_Report.bib
---

@@

#setwd("~./MLH1repo/doc/")
#setwd("C:/Users/alpeterson7/Documents/MLH1repo/doc/")
#setwd("C:/Users/April/Documents/MLH1repo")
#load(file="data/MLH1/MLH1_data_setup.RData")

#load main data file
#load(file="C:/Users/April/Documents/MLH1repo/data/MLH1/MLH1_data_setup_1.5.20.RData")#added batch18
#load(file="C:/Users/April/Documents/MLH1repo/data/CleanBivData_1.28.20.RData")
#load(file="C:/Users/alpeterson7/Documents/MLH1repo/data/CleanBivData_1.28.20.RData")

#load(file="data/MetaData.RData") #this was replacing most updated data file

@@

```{r table.for.plot, echo=FALSE, warning=FALSE}
#why do I add in a different table?

#this move this to the setup file
#MLH1_data  %>% filter(quality < 5)  -- low quality cells removed

@@

```{r Fig1, echo=FALSE, warning=FALSE}

#might need to nudge the X values

@@

    theme(legend.position="none",
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"),
  axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank())
#remove y labfor male

@@

female.specific.plot <- ggplot(female_mouse.avs_for.analysis, aes(x = as.factor(strain), y = mean_co, color= strain))+ geom_boxplot()+  geom_jitter(aes( alpha=0.5), width = .2 )+ ggtitle("Female mouse av")+scale_color_manual(values=colors_of_strains)+ylim(c(20,34))+
  theme(legend.position="none",
      axis.text.x=element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"))

@@

male_mouse.av <- subset(AP_mouse_table_w.Ages, select= -c(sex.y,sex.x,
                            subsp.y,subsp.x,
                    category.y,category.x, strain.x,
                    strain.y, mouse.1  )  )

#just.male_mouse.av <-  male_mouse.av %>% filter(sex== "male") %>% #filter(species == "M.musculus") 
#just.male_mouse.av <-  fomo.male_mouse.av %>% filter(sex== "male") %>% filter(species == "M.musculus")

#just.male_mouse.av$sex <- factor(just.male_mouse.av$sex, ordered = FALSE)
#just.male_mouse.av$strain <- factor(just.male_mouse.av$strain, ordered = FALSE)
#just.male_mouse.av$subsp <- factor(just.male_mouse.av$subsp, ordered = FALSE)
#just.male_mouse.av$species <- factor(just.male_mouse.av$species, ordered = FALSE)

#table( droplevels(just.male_mouse.av$strain) )
#there are cast strains too
#Musc.male_mouse.av <-  just.male_mouse.av %>% filter(sex== "male") %>% filter(subsp == "Musc")
#fomo.male_mouse.av

male.plot <- ggplot(just.male_mouse.av, aes(x = as.factor(strain), y = mean_co, color= strain))+geom_jitter( alpha=0.5, width = .2) +geom_boxplot()+ ggtitle("Male mouse mean")+scale_color_manual(values=colors_of_strains)+ylim(c(20,34))+
  theme(legend.position="none", axis.text.x=element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), 
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"))

```{r three.plots, echo=FALSE, warning=FALSE}
#merge the 3 plots
#try patchwork

#Fig1_v1_strain + female.specific.plot + male.plot

#Fig1_v1_strain + female.specific.plot - male.plot + plot_layout(ncol=1)
Fig1_v1_strain / (female.specific.plot + male.plot )

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice: *M. m. domesticus*, *M. m. musculus* and *M. m. molossinus*. Mean MLH1 foci counts for `r length( unique(AP_mouse_table.HQ %>% filter(species == "M.musculus") )$mouse )` mice were quantified from an average of `r mean( (male_mouse.av %>% filter(species == "M.musculus") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "male"))$Ncells)` spermatocytes) and `r mean(female_mouse.avs_for.analysis$Ncells, na.rm = T)` oocytes per female (for a total of `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells)` oocytes).

@@

#reorder the df.strain.table
#species, subspecies, strain

#code for reording column names...
#DMC1table.main99 <-  DMC1table.main99[,c("category","mean.MLH1","ncells.x","mean.count.L", "MLH1.L_ratio", #"ncells.y","mean.count.Z", "MLH1.Z_ratio"),  drop=FALSE]

#write.xlsx(df.strain.table, file = "Results_Table.xlsx", sheetName = "MLH1.results", append = T, row.names = F)

@@

To situate variation in recombination rate within an evolutionary framework, we fit a series of models including subspecies, strain, and sex, to mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice. We began with a full mixed model (M1, see Methods), which showed that strain 
(random effect p < `r MLH1.MMlmer_results$rand.pval` -- `r 1/1000000` someSmallNumber), sex (p = `r MLH1.MMlmer_results$sex.results`), subspecies (p=`r MLH1.MMlmer_results$sub.sp`), and subspecies sex (p = `r MLH1.MMlmer_results$interaction.pval`) each significantly affect recombination rate.

After a general linear model including all factors as fixed effects (M2) revealed only weak contributions of subspecies, we focused on additional models designed to illuminate the role of strain and sex. A general linear model with these two variables (M3) identified two strains with particularly strong effects on recombination rate: musculus^MSM^ (p= `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`)
and domesticus^G^ (`r summ.MLH1.M3.strain.fixed$coefficients[,4][3]`). In addition, two strains exhibit strain-by-sex interactions:  molossinus^MSM^  (`r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`) and musculus^PWD^ (`r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`).

@@

Counting MLH1 foci in multiple oocytes for each female and multiple spermatocytes for each male allowed us to examine determinants of the within-mouse variance in recombination rate. To do this, we considered the same models as above, but replaced mean MLH1 foci count with within-mouse variance in MLH1 foci count as the dependent variable. Sex is the only variable that significantly affects recombination rate in both the mixed model (M1) (p < `r 1/1000000` -- `r round(LmerMLH1_VAR_results$sex.results, 5)`) and general linear models (M2) (p = `r MLH1var_M2_sum$coefficients[,4][9]`) and M3 (p = `r round(glmMLH1Var_M3_sum$coefficients[,4][2], 5)`).

In general, females have almost twice as much variance in MLH1 foci per cell compared to males (Figure 1). Since estimates of within-mouse variance may be more susceptible to technical error from the staining protocol, we repeated the analyses using a subset of cells with higher quality scores (quality score 1 and 2, see Materials and Methods).  The results are similar: sex is the strongest effect (M1 p < `r 1/1000000` `r round(LmerMLH1_M1_Q12_VAR_results$sex.results, 5)`; M2 p = `r  round(sum.MLH1var_M2_Q12$coefficients[,4][4], 5)`;  also M3 p = `r MLH1var_M3_Q12_sum$coefficients[,4][2]`). When both quality-curated and full datasets are considered, strain does not significantly and consistently affect variance in MLH1 foci count per cell in either sex. These results suggest that within-mouse variance in recombination rate evolves independently of mean recombination rate.

@@

```{r MAIN.DMC1.plot, echo=FALSE, warning=FALSE}

@@

#change order

DMC1.DF$strain<- factor(DMC1.DF$strain, levels =c("LEW",
                      "WSB","G", "KAZ", "SKIVE","PWD","MSM"), order=T )

DMC1.boxplot <- ggplot(aes(x=strain, y=total.count, color=stage), data=DMC1.DF[DMC1.DF$stage != "P",])+
  geom_boxplot(alpha=.5) +geom_jitter()+ggtitle("DMC1 averages in juvenile mice\nZygotene spermatogcytes")+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

DMC1.boxplot
```

@@

In an attempt to localize the male-specific evolution of crossover number to steps of the meiotic pathway, we counted foci from a marker for double strand breaks (DSBs), DMC1, in prophase spermatocytes. DMC1 foci were scored from a total of `r sum( (DMC1.mean.table %>%  filter(stage == "Early") )$ncells)` early zygotene-stage and `r sum( (DMC1.mean.table %>%  filter(stage == "Late") )$ncells)` late zygotene-stage spermatocytes from three low-recombination strains (musculus^KAZ^ , domesticus^WSB^ , and domesticus^G^) and two high-recombination strains (musculus^PWD^ and molossinus^MSM^).

The high-recombination strains have significantly more DMC1 foci than the low-recombination strains in early zygotene cells (t-test, p<  -- `r 1/1000000` -- `r ttest.HighLow.L.pval`). In contrast, the two strain groups do not differ in DMC1 foci counted in late zygotene cells (t-test, p = 0.66 --- `r ttest.HighLow.Z.pval`).

After DSB formation, DSBs are repaired as either non-crossovers (NCO) or as crossovers (COs), with the vast majority being repaired as NCOs. Thus the ratio of CO:DSB is a partial indicator of the proportion of DSBs which are designated as COs. The ratios, calculated for DMC1 means from both stages, are not significantly different between the high and low strain groups (t-test, p = `r DMC1.Lratio_t.test$p.value` and p = `r DMC1.Zratio_t.test$p.value` for early zygotene and late zygotene ratios, respectively).

This comparison raises the possibility that the evolution of crossover number is primarily due to processes that precede the crossover/non-crossover decision. [@cole2012] suggest that early zygotene counts for DMC1 are most relevant for predicting the final CO number due to the crossover homeostasis process. This result, combined with the big difference in MLH1 count between high and low strains, might indicate that the CO/NCO decision contributes to strain differences in recombination rate.

@@

<bairer-- DSB number is not a primary driver of variation in CO number -- these authors suggest that chromatin organizaiton (loop-axis) structure is the primary determinant of CO variation.>

@@

```{r Chrm.props.plot2, echo=FALSE}

@@

chrm.class.barplot <- ggplot(data = chrm.class.category_MELT, aes(y = mean.percent.type, x = strain, fill=as.factor(variable)) ) + geom_col(position = "dodge")+ggtitle("Chrm Proportions from BivData Data")+ylim(c(0,1)) + facet_wrap(sex~subsp, scales = "free")
#chrm.class.barplot
chrm.class.barplot

@@

From the total set of cell images, `r length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName )` bivalent objects were isolated by the image analysis software. After a human curation step ([@peterson2019]), `r length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID )`  single-bivalent observations remained. We assume that the isolation of bivalents within cells is unbiased. <might be the only section to leave in Results>

Given the large number of single-bivalent observations, we assume that each of the datasets are equally representative of general patterns. <Bret says this statment should be more clear>

An additional challenge of the MLH1 framework is that the identities of individual autosomes and the XX in females cannot be easily obtained (the male XY is distinct). <Bret suggests moving to Methods>

Ninety-six percent of single bivalents in our pooled dataset (n = `r total.biv.all`) have either one or two crossovers (Figure X). The proportion of one-crossover (1CO) to two-crossover (2CO) bivalents distinguishes the high vs. low recombining strains (Figure X). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are 0.33 -- `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` ( _musculus^SKIVE^_ ) in musculus^SKIVE^, 0.44 -- `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` in musculus^PWD^, and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` in molossinus^MSM^.

@@

#CLEAN UP THE FULL DATA SET

@@

#nested boxplots, pooled Bivalents

Q1.SC.length_Nested_boxplots_Dom <- ggplot(Curated_BivData99[Curated_BivData99$subsp == "Dom",], aes(y = chromosomeLength, x=strain, fill=as.factor(hand.foci.count) ))+geom_boxplot(alpha=0.5)+ggtitle("Bivalent SC Lengths by Chrm Class \n Dom")+facet_wrap(subsp~sex, scales="free")+ylim(c(25,255))+theme(legend.position="none")

Q1.SC.length_Nested_boxplots_Musc <- ggplot(Curated_BivData99 %>% filter(subsp == "Musc") %>% filter(strain != "CZECH"), aes(y = chromosomeLength, x=strain, fill=as.factor(hand.foci.count) ))+geom_boxplot(alpha=0.5)+ggtitle("Bivalent SC Lengths by Chrm Class \nMusc")+facet_wrap(subsp~sex, scales="free")+ylim(c(25,255))+theme(legend.position="none")

Q1.SC.length_Nested_boxplots_Mol <- ggplot(Curated_BivData99 %>% filter(strain == "MSM"), aes(y = chromosomeLength, x=as.factor(hand.foci.count), fill=as.factor(hand.foci.count) ))+geom_boxplot(alpha=0.5)+geom_jitter(alpha=.3)+
  ggtitle("Bivalent SC Lengths by Chrm Class\n Mol")+facet_wrap(strain~sex, scales="free")+ylim(c(25,255))+theme(legend.position="none")

@@

Q1.SC.short.mouse_scatter_boxplot <- ggplot( Q1.real.short_mouse_table, aes(y = mean_SC, x = sex, color=strain)) + geom_jitter()+ geom_boxplot(alpha=.2) +scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales="free")+ylim(c(20,115) )+ggtitle("Mouse average SC Length \n short bivalent dataset")

Q1.SC.short.mouse_scatter_boxplot

#ploting SKIVE alone, confirms the ranges overlap, KAZ also have close ranges
skiveQ1.SC.short.mouse_scatter_boxplot <- ggplot( Q1.real.short_mouse_table %>% filter(strain == "SKIVE"), aes(y = mean_SC, x = sex, color=strain)) + geom_jitter()+ geom_boxplot(alpha=.2) +scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales="free")+ylim(c(20,115) )+ggtitle("Mouse average SC Length \n short bivalent dataset")
#skiveQ1.SC.short.mouse_scatter_boxplot
#there are 2 lowish female means,

@@

```{r Q1SC_short.biv_dif, echo=FALSE}

#calq the different in short.biv means

@@

```{r Q1.tot.SC_show, echo=FALSE, warning=FALSE}
# fig.width = 4, fig.height=2.5

@@

#COPY THIS CODE TO BELOW
#Q1.SC.long.mouse_scatter_boxplot <- ggplot( Q1.real.long_mouse_table, aes(y = mean_SC, x = sex, color=strain)) + geom_jitter()+ geom_boxplot(alpha=.2) +scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales="free")+ylim(c(75,150) )+ggtitle("Mouse average SC Length \n long bivalent dataset")

#Q1.SC.short.mouse_scatter_boxplot <- ggplot( Q1.real.short_mouse_table, aes(y = mean_SC, x = sex, color=strain)) + geom_jitter()+ geom_boxplot(alpha=.2) +scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales="free")+ylim(c(20,115) )+ggtitle("Mouse average SC Length \n short bivalent dataset")

@@

  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+  ylim(c(1000,2600 ))+ggtitle("Mouse av. total SC area") + 
  #change the white themes
  theme(legend.position="none")

Tot.SC_mouse.av_ALL

totsc.musc <- Q1.totSC.mouse.av.table %>% filter( subsp == "Musc" )
totsc_strain.av.msc <- totSC.strain.av.table %>% filter( subsp == "Musc" )

Tot.SC_mouse.av_Musc <- ggplot(data= totsc.musc, aes(y = mean.skel, x=mouse, color=strain))+ geom_point()+scale_color_manual(values=colors_of_strains)+
  facet_wrap(~sex, scales="free", ncol = 3)+

  geom_hline(data =totsc_strain.av.msc, aes(yintercept=mean.skel) )+

  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+  ggtitle("Mouse av. total SC area")+  ylim(c(1000,3000 ))

totsc.mol <- Q1.totSC.mouse.av.table %>% filter( subsp == "Mol" )
totsc_strain.av.mol <- totSC.strain.av.table %>% filter( subsp == "Mol" )

Tot.SC_mouse.av_Mol <- ggplot(data= totsc.mol, aes(y = mean.skel, x=mouse, color=strain))+ geom_point()+scale_color_manual(values=colors_of_strains)+
  facet_wrap(~sex, scales="free", ncol = 3)+

  geom_hline(data =totsc_strain.av.mol, aes(yintercept=mean.skel) )+

  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+  ylim(c(1000,3000 ))+ggtitle("Mouse av. total SC area")

@@

In many mammalian species, the synaptonemal complex (SC) is longer in females. Yet, the majority of these observations come from species with female-biased heterochiasmy. Our wider survey of recombination in house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is male-biased. In addition, if SC length is a strong determinant of the genome-wide recombination rate,  male musculus^PWD^ and male molossinus^MSM^ should have longer SC lengths than the other strains. <moved section on reduced bivalent data set to methods>

@@

Short bivalents are significantly longer in females than males in all strains (t-test; p < 0.05) except musculus^SKIVE^.  The female:male ratio ranges across strains from `r Q1SC.mus.av.short.biv_DIFF_MSM` (musculus^MSM^) to `r Q1SC.mus.av.short.biv_DIFF_WSB` (domesticus^WSB^). <moved section on total SC to methods>

Females have significantly longer total SC than males in each strain tested separately (t-test; p < 0.05). That females have longer SCs is also supported by mixed models and general linear models with covariates, which identify sex as the most consistently significant effect (p < 0.05). Additionally, there are some significant subspecies and strain effects (p < 0.05), indicating that SC length has evolved among strains and subspecies.

@@

```{r double.tri.rewrite, echo=FALSE}

@@

>>>>

---
title: "MLH1 Results Outline"
bibliography: refs_Main_Report.bib
date: 'null'
output:
  word_document:
    toc: yes
    toc_depth: '5'
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
always_allow_html: yes
---

@@

@@

```{r table.for.plot, echo=FALSE, warning=FALSE}
#why do I add in a different table?
#this move this to the setup file
#MLH1_data  %>% filter(quality < 5)  -- low quality cells removed

@@

```{r Fig1.top, echo=FALSE, warning=FALSE}

@@

    theme(legend.position="none",
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"),
  axis.title.y=element_blank(),
  axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank())
#remove y labfor male

@@

female.specific.plot <- ggplot(female_mouse.avs_for.analysis, aes(x = as.factor(strain), y = mean_co, color= strain))+ geom_boxplot()+  geom_jitter(aes( alpha=0.5), width = .2 )+ ggtitle("")+scale_color_manual(values=colors_of_strains)+ylim(c(20,34))+
  theme(legend.position="none",

        axis.title.y =element_blank(), 
      axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"))

@@

male.plot <- ggplot(just.male_mouse.av, aes(x = as.factor(strain), y = mean_co, color= strain))+geom_jitter( alpha=0.5, width = .2) +geom_boxplot()+ ggtitle("")+scale_color_manual(values=colors_of_strains)+ylim(c(20,34))+

   theme(legend.position="none", axis.text.x=element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(),  axis.title.y = element_blank(), 
  panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line=element_line(colour = "black"))

```{r Figure1.Main.plot.three.plots, echo=FALSE, warning=FALSE, fig.width=7}

Figure1.main.plot <- Fig1_v1_strain / (female.specific.plot + male.plot )

Figure1.main.plot + plot_annotation(
  title = '',
  subtitle = '',
  caption = 'caption'
) + plot_annotation(tag_levels = 'A')

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice: *M. m. domesticus*, *M. m. musculus* and *M. m. molossinus*. Mean MLH1 foci counts for `r length( unique(AP_mouse_table.HQ_target)$mouse )` mice were quantified from an average of `r mean( (AP_mouse_table.HQ_target %>% filter(sex == "male") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "male"))$Ncells),  format="d", big.mark=",")` spermatocytes) and `r mean(female_mouse.avs_for.analysis$Ncells, na.rm = T)` oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` oocytes).

@@

@@

To situate variation in recombination rate within an evolutionary framework, we fit a series of models including subspecies, strain, and sex, to mean MLH1 foci counts from `r length(DF.HetC.MixedModel.HQ$mean_co)` mice. We began with a full mixed model (M1, see Methods), which showed that strain 
(random effect p < <r MLH1.MMlmer_results\$rand.pval> `r 1/1000000`), sex (p = `r MLH1.MMlmer_results$sex.results`), subspecies (p=`r MLH1.MMlmer_results$sub.sp`), and subspecies sex (p = `r MLH1.MMlmer_results$interaction.pval`) each significantly affect recombination rate.

After a general linear model including all factors as fixed effects (M2) revealed only weak contributions of subspecies, we focused on additional models designed to illuminate the role of strain and sex. A general linear model with these two variables (M3) identified two strains with particularly strong effects on recombination rate: musculus^MSM^ (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`)
and domesticus^G^ (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]`). In addition, two strains exhibit strain-by-sex interactions:  molossinus^MSM^  (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`) and musculus^PWD^ (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`).

@@

Counting MLH1 foci in multiple oocytes for each female and multiple spermatocytes for each male allowed us to examine determinants of the within-mouse variance in recombination rate. To do this, we considered the same models as above, but replaced mean MLH1 foci count with within-mouse variance in MLH1 foci count per cell as the dependent variable. Sex is the only variable that significantly affects recombination rate in both the mixed model (M1) (p < `r 1/1000000`) and general linear models (M2) (p = `r MLH1var_M2_sum$coefficients[,4][9]`) and M3 (p = `r round(glmMLH1Var_M3_sum$coefficients[,4][2], 5)`).

In general, females have almost twice as much variance in MLH1 foci per cell compared to males (Figure 1). Since estimates of within-mouse variance may be more susceptible to technical error from the staining protocol, we repeated the analyses using a subset of cells with higher quality scores (quality score 1 and 2, see Materials and Methods).  The results are similar: sex is the strongest effect (M1 p < `r 1/1000000` <round(LmerMLH1_M1_Q12_VAR_results\$sex.results, 5)>; M2 p = `r  round(sum.MLH1var_M2_Q12$coefficients[,4][4], 5)`;  also M3 p = `r MLH1var_M3_Q12_sum$coefficients[,4][2]`). When both quality-curated and full datasets are considered, strain does not significantly and consistently affect variance in MLH1 foci count per cell in either sex. These results suggest that within-mouse variance in recombination rate evolves independently of mean recombination rate.

**(sex specific results - no efficient strain effect for males. In females LEW - but this is due to cell quality difference, what about quality score and cell number within the models?**

@@

```{r DMC1.boxplot.plot, echo=FALSE, warning=FALSE}

@@

#change order
DMC1.DF$strain<- factor(DMC1.DF$strain, levels =c("LEW",
                      "WSB","G", "KAZ", "SKIVE","PWD","MSM"), order=T )

DMC1.boxplot <- ggplot(aes(x=strain, y=total.count, color=stage), data=DMC1.DF[DMC1.DF$stage != "P",])+
  geom_boxplot(alpha=.5) +geom_jitter() +

    labs(y = "DMC1 count per Cell")+

  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.title.x = element_blank())

```{r Main.plot.DMC1, echo=FALSE, fig.width=6.5, fig.height=5}

png.DMC1.early.cell <- readPNG('~/MLH1repo/doc/figureFiles/cell.examples/DMC1.early/good.early.cell.png')
png.DMC1.late.cell <- readPNG('~/MLH1repo/doc/figureFiles/cell.examples/DMC1.late/good.late.cell.png')

empty.DF <- data.frame(col1 = c(1,3), col2 = c(1,3) )
#EMPTY GGPLOT
Early.cell.obj <- ggplot(empty.DF, aes(x = col1, y = col2)) +
  background_image(png.DMC1.early.cell) + theme(axis.title.x=element_blank(),
                         axis.text.x=element_blank(),axis.ticks.x=element_blank(),axis.title.y=element_blank(),
                                  axis.text.y=element_blank(),axis.ticks.y=element_blank() )

Late.cell.obj <- ggplot(empty.DF, aes(x = col1, y = col2)) +
  background_image(png.DMC1.late.cell) + theme(axis.title.x=element_blank(),
                         axis.text.x=element_blank(),axis.ticks.x=element_blank(),axis.title.y=element_blank(),
                                  axis.text.y=element_blank(),axis.ticks.y=element_blank() )

three.DMC1.plot <- { Early.cell.obj + Late.cell.obj  } / DMC1.boxplot + plot_layout(widths = 1)

#+ plot_layout(ncol = 2, widths = c(.65, .65), heights = c(.65, .65))
#plot_layout(nrow = 2, widths = c(1), heights = c(.45) )

#plot_grid(mouse.av_totsc_male, X.by.texas, nrow=1, labels=c('A', 'B')) #Or #labels="AUTO"

three.DMC1.plot + plot_annotation(
  title = '',
  subtitle = '',
  caption = 'caption'
) + plot_annotation(tag_levels = 'A')

@@

In an attempt to localize the male-specific evolution of crossover number to steps of the meiotic pathway, we counted foci from a marker for double strand breaks (DSBs), DMC1, in prophase spermatocytes. DMC1 foci were scored from a total of `r sum( (DMC1table.main99$ncells) )` early zygotene-stage and `r sum( (DMC1table.main99$ncells) )` late zygotene-stage spermatocytes from three low-recombination strains (musculus^KAZ^ , domesticus^WSB^ , and domesticus^G^) and two high-recombination strains (musculus^PWD^ and molossinus^MSM^).

The high-recombination strains have significantly more DMC1 foci than the low-recombination strains in early zygotene cells (t-test, p < `r 1/1000000` <ttest.HighLow.L.pval>). In contrast, the two strain groups do not differ in DMC1 foci counted in late zygotene cells (t-test, p = `r ttest.HighLow.Z.pval`).

After DSB formation, DSBs are repaired as either non-crossovers (NCO) or as crossovers (COs), with the vast majority being repaired as NCOs. Thus the ratio of CO:DSB is a partial indicator of the proportion of DSBs which are designated as COs. The ratios, calculated for DMC1 means from both stages, are not significantly different between the high and low strain groups (t-test, p = `r DMC1.Lratio_t.test$p.value` and p = `r DMC1.Zratio_t.test$p.value` for early zygotene and late zygotene ratios, respectively). (connection to barier)

This comparison raises the possibility that the evolution of crossover number is primarily due to processes that precede the crossover/non-crossover decision. @cole2012 suggest that early zygotene counts for DMC1 are most relevant for predicting the final CO number due to the crossover homeostasis process.

This result, combined with the large difference in MLH1 count between high and low strains, might indicate that the CO/NCO decision contributes to strain differences in recombination rate.

@@

@@

```{r Chrm.props.plot2, echo=FALSE, fig.width=8}

@@

chrm.class.barplot <- ggplot(data = chrm.class.category_MELT, aes(y = mean.percent.type, x = strain, fill=as.factor(variable)) ) + geom_col(position = "dodge")+ggtitle("")+ylim(c(0,1)) + facet_wrap(sex~subsp, scales = "free")+ theme(legend.position="none", axis.title.y = element_blank(), axis.text.x = element_text(size=9), axis.title.x = element_blank() )
#chrm.class.barplot
chrm.class.barplot

@@

From the total set of cell images, `r formatC(sum( length(unique(new.Curate.FULL.org %>% filter(SC.pass == "1"))$fileName ) ), , format="d", big.mark="," )` bivalent objects were isolated by the image analysis software. After a human curation step ([@peterson2019]), `r formatC(sum(  length(unique(new.Curate.FULL.org %>% filter(hand.foci.count != ""))$Obj.ID ) ), , format="d", big.mark=",")`  single-bivalent observations remained. We assume that the isolation of bivalents within cells is unbiased.

Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( total.biv.all, format="d", big.mark=",")`) have either one or two crossovers (Figure X). The proportion of one-crossover (1CO) to two-crossover (2CO) bivalents distinguishes the high vs. low recombining strains (Figure X). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` ( _musculus^SKIVE^_ ) in musculus^SKIVE^, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` in musculus^PWD^, and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` in molossinus^MSM^.

@@

@@

#nested boxplots, pooled Bivalents
Q1.SC.length_Nested_boxplots_Dom <- ggplot(Curated_BivData99[Curated_BivData99$subsp == "Dom",], aes(y = chromosomeLength, x=strain, fill=as.factor(hand.foci.count) ))+geom_boxplot(alpha=0.5)+ggtitle("Bivalent SC Lengths by Chrm Class \n Dom")+facet_wrap(subsp~sex, scales="free")+ylim(c(25,255))+theme(legend.position="none")

@@

Q1.SC.short.mouse_scatter_boxplot <- ggplot( Q1.real.short_mouse_table, aes(y = mean_SC, x = sex, color=strain)) + geom_jitter()+ geom_boxplot(alpha=.2) +scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales="free")+ylim(c(20,115) )+ggtitle("")+  theme(legend.position="none", axis.title.x = element_blank(), axis.text.x = element_text(size=9), axis.text.y = element_text(size=9))+labs(y = "Mouse Mean Short SC Length")

@@

```{r Q1SC_short.biv_dif, echo=FALSE}
#calq the different in short.biv means across sexes

@@

```{r Q1.tot.SC.plot, echo=FALSE, warning=FALSE}
# fig.width = 4, fig.height=2.5

@@

@@

  theme(axis.title.x=element_blank(), axis.ticks.x=element_blank(),axis.text.x = element_text(size=9), legend.position="none", axis.text.y = element_text(size=9))+  
  labs(y = "Mouse Mean Short SC Length")+
  ylim(c(1000,2600 ))+ggtitle("")

```{r Q1.Two.SC.plots, echo=FALSE, warning=FALSE, fig.width=8}

two.Q1.SC.plots <- {Q1.SC.short.mouse_scatter_boxplot + {Tot.SC_mouse.av_ALL} + plot_layout(ncol = 2, widths = c(1,1), heights = c(1,1)) }

two.Q1.SC.plots + plot_annotation(
  title = '',
  subtitle = '',
  caption = 'caption'
) + plot_annotation(tag_levels = 'A')

@@

In many mammalian species, the synaptonemal complex (SC) is longer in females (@cahoonLibuda2019, @sardell_sex_2020). Yet, the majority of these observations come from species with female-biased heterochiasmy. Our wider survey of recombination in house mice provides an opportunity to determine whether sex differences in chromatin compaction (SC length) are reversed when heterochiasmy is male-biased. In addition, if SC length is a strong determinant of the genome-wide recombination rate,  male musculus^PWD^ and male molossinus^MSM^ should have longer SC lengths than the other strains. <moved section on reduced bivalent data set to methods>

@@

We use the reduce data set of short bivalents to remove the effects of the paired X-X bivalent in females. The mouse average of short bivalents are significantly longer in females than males in all strains (t-test; p < 0.05) except musculus^SKIVE^, which my be attributed to low sample sizes (p = `r Q1_mus.av_shortSC_t.test.SKIVE$p.value`). The ratios of the mouse mean SC across sexes range across strains from `r Q1SC.mus.av.short.biv_DIFF_MSM` (musculus^MSM^) to `r Q1SC.mus.av.short.biv_DIFF_WSB` (domesticus^WSB^).

Females have significantly longer total SC than males in each strain tested separately (t-test; p < 0.05). That females have longer SCs is also supported by mixed models and general linear models with covariates, which identify sex as the most consistently significant effect (p < 0.05). Additionally, there are some significant subspecies and strain effects (p < 0.05), indicating that SC length has evolved among strains and subspecies, however further investigation is beyond the scope of this paper.

@@

```{r double.IFD.triangle.plots, echo=FALSE}

@@

x.dist <- .1
RP <- list()
